Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-09-17 writeup/pnas-word-order-si.tex
code

@@

	\item Correlation \multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}} is formalized using \emph{nmod}, which covers not only PPs and standards of comparison, but also adjunct NPs.
		In the WALS evaluation (Table~\ref{tab:grammars-wals}), the dominant order of \emph{nmod} agrees with that annotated for obliques in all 18 languages for which WALS has an entry.

@@

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage-raw.pdf}
	\caption[Predictability and Parseability]{Raw numerical values estimated for Predictability (negative surprisal), and negative syntactic ambiguity $-H[T|U]$, before $z$-scoring. For more meaningful comparison, both quantities are normalized by the number of words in the corpus, i.e., we plot per-word negative surprisal and per-word difficulty in recovering syntactic structures.  Note that the negative syntactic ambiguity $-H[T|U]$ equals parseability $I[T,U] = H[T] - H[T|U]$ up to a per-language constant $H[T]$, which we do not attempt to estimate.}\label{fig:pareto-per-lang-raw}
\end{figure}

@@

@@

We eliminate this anti-conservative bias by conducting joint Bayesian inference of $\lambda$ of $O$, using the Dirichlet Gaussian Process, a Bayesian version of Kernel Density Smoothing.
The 50 samples represent an underlying baseline distribution in the predictability-parseability plane.
We approximate this underlying distribution as a continuous distribution from the Dirichlet Gaussian Process, and infer its parameters using Bayesian inference.
We use this method to obtain a Bayesian estimate of the distribution of baseline distributions in the Surprisal-Parseability plane.
The Dirichlet Gaussian Process (CITE) defines a prior over possible baseline distributions.
Under this prior, each possible distribution is parameterized by a number $n$ of components, and $n$ Gaussians $\mu_n, \sigma^2_n$.
Due to the flexible number of components, this nonparametric prior can accomodate a wide range of continuous probability distributions, including multimodal ones.
We use the $50$ observed baseline samples to infer a posterior distribution over such smooth distributions.
We then use MCMC to obtain 1,000 samples from the posterior, based on these samples.
We used the MCMC implementation provided by the \texttt{R} package \texttt{dirichletprocess}~\citep{ross2018dirichletprocess}, discarding the first 1,000 samples as warmup samples, and obtaining another 1,000 samples.

@@

We quantify the Pareto-optimality of the language $r$ as the fraction of baselines that are better than $r$ on \emph{both} parseability and predictability:
\begin{equation}
	PO := \frac{\# \{b \in B: R_{Pars}(b) > R_{eff}(r) \text{ and }  R_{Pred}(b) > R_{Pred}(r) \}}{\#B}
\end{equation}
This quantifies how many hypothetical grammars are more efficient than $r$ \emph{irrespective} of $\lambda$.

>>>>

TODO - say that, beyond just what Dryer reported, we see this distribution in the UD languages, and reproduce this

@@

	\item Correlation \multirow{2}{*}{\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}} is formalized using \emph{obl}, which covers not only PPs and standards of comparison, but also adjunct NPs.
		In the WALS evaluation (Table~\ref{tab:grammars-wals}), the dominant order of \emph{nmod} agrees with that annotated for obliques in all 18 languages for which WALS has an entry.

@@

% say this somewhere
Note that the predictability scores are considerably worse than the state-of-the art in Natural Language Processing; this is because most of the corpora are much smaller than those typically used in language modeling work.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage-raw.pdf}
	\caption[Predictability and Parseability]{Raw numerical values estimated for Predictability (negative surprisal), and negative syntactic ambiguity $-H[T|U]$, before $z$-scoring. For more meaningful comparison, both quantities are normalized by the number of words in the corpus, i.e., we plot per-word negative surprisal and per-word difficulty in recovering syntactic structures.  Note that the negative syntactic ambiguity $-H[T|U]$ equals parseability $I[T,U] = H[T] - H[T|U]$ up to a per-language constant $H[T]$, which we do not attempt to estimate. Further note that we use different scales in the different panels. }\label{fig:pareto-per-lang-raw}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/analyze_pareto_optimality/figures/quantileByLambda.pdf}
	\caption[]{The $x$-axis shows $\lambda \in [0,1)$, the $y$-axis shows the fraction of baselines that have lower efficiency at this value of $\lambda$, with 95\% confidence bands obtained from a binomial test.}\label{fig:lambda-quantile}
\end{figure}

\paragraph{Optimality as a Function of $\lambda$}

Figure~\ref{fig:lambda-quantile}

@@

We eliminate this anti-conservative bias by conducting joint Bayesian inference of $\lambda$ of $O$, using the Dirichlet Gaussian Process, a Bayesian version of Kernel Density Smoothing.
The 50 samples represent an underlying baseline distribution in the predictability-parseability plane.
We approximate this underlying distribution as a continuous distribution from the Dirichlet Gaussian Process, and infer its parameters using Bayesian inference.
We use this method to obtain a Bayesian estimate of the distribution of baseline distributions in the Surprisal-Parseability plane.
The Dirichlet Gaussian Process (CITE) defines a prior over possible baseline distributions.
Under this prior, each possible distribution is parameterized by a number $n$ of components, and $n$ Gaussians $\mu_n, \sigma^2_n$.
Due to the flexible number of components, this nonparametric prior can accomodate a wide range of continuous probability distributions, including multimodal ones.
We use the $50$ observed baseline samples to infer a posterior distribution over such smooth distributions.
We then use MCMC to obtain 1,000 samples from the posterior, based on these samples.
We used the MCMC implementation provided by the \texttt{R} \citep{R2013} package \texttt{dirichletprocess}~\citep{ross2018dirichletprocess}, discarding the first 1,000 samples as warmup samples, and obtaining another 1,000 samples.

@@

We quantify the Pareto-optimality of the language $r$ as the fraction of baselines that are better than $r$ on \emph{both} parseability and predictability:
\begin{equation}
	PO := \frac{\# \{b \in B: R_{Pars}(b) > R_{eff}(r) \text{ and }  R_{Pred}(b) > R_{Pred}(r) \}}{\#B}
\end{equation}
This quantifies how many hypothetical grammars are more efficient than $r$ for \emph{all} choices of $\lambda$.
